---
title: "Lab 4: INFORMATIVE TITLE"
subtitle: "w203: Statistics for Data Science, Section 2"
author: "Krissy Gianforte & Dan Kent"
date: "18 December 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE}
library(car)
library(sandwich)
library(stargazer)
library(lmtest)
```

# Introduction

G&K Associates have been retained by the [REDACTED] campaign to provide a statistical modeling and analysis to understand the determinants of crime and generate policy suggestions that are applicable to local government.  We, the principal investigators (K. Gianforte & D. Kent) utilized a pre-existing dataset of crime statistics for a selection of counties and are proceeding in this exercise from a causal perspective.  


# Initial Exploratory Analysis
```{r}
setwd("C:\\Users\\Krissy\\Dropbox\\W203 - Statistics\\Lab 4")
data <- read.csv("crime_v2_updated.csv", header = TRUE)

summary(data$crime)
hist(data$crime)

```  

# Model Building
## Model 1: 
#### Specification

#### Assumptions

## Model 2:  
#### Specification

#### Assumptions

## Model 3:  
#### Specification

#### Assumptions

## Summary of Models

# Summary of models
## Regression Table

## Significance
### Statistical Significance

### Practical Significance



# Causality & Omitted Variables

# Conclusion
