---
title: "Lab 4: INFORMATIVE TITLE"
subtitle: "w203: Statistics for Data Science, Section 2"
author: "Krissy Gianforte & Dan Kent"
date: "18 December 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE}
library(car)
library(sandwich)
library(stargazer)
library(lmtest)
```

# Introduction

G&K Associates have been retained by the [REDACTED] campaign to provide a statistical modeling and analysis to understand the determinants of crime and generate policy suggestions that are applicable to local government.  We, the principal investigators (K. Gianforte & D. Kent) utilized a pre-existing dataset of crime statistics for a selection of counties and are proceeding in this exercise from a causal perspective.  


# Initial Exploratory Analysis
```{r}
setwd("C:\\Users\\Krissy\\Dropbox\\W203 - Statistics\\Lab 4")
data <- read.csv("crime_v2_updated.csv", header = TRUE)
data <- data[2:26]
```  

We begin our Exploratory Data Analysis by first inspecting all the variables.  We identify 25 different variables and confirm that these are the variables provided to us in the supplemental codebook.

```{r}
summary(data)
```
From our high-level summary statistics and descriptions from the codebook, we observe that some variables are categorial-ordinal (county, year), coded (west, central, urban), proportions or probabilities (probarr, probconv, probsen, pctmin, mix, ymale), averages (crime, avgsen, wagecon, wagetuc, wagetrd, wagefir, wageser, wagemfg, wagefed, wagesta, wageloc), and some are rates (crime, police, density, tax).  

We first identify our response/dependent variable, crime, the quantity of "crimes committed per person."

We can initally facially exclude a series of variables based upon the following conditions:

year - The year variable can be ignored in our analysis as all the observations have the same value, 88, which the researchers understand as 1988 - perhaps the year of the data.  As all of the values are the same across all observations, we ignore this variable.

probarr, probconv, probsen - The variables involving probability will be excluded from analysis as they ______

mix - the mix variable, described in the codebook as "ratio of face to face/all other crimes" will be excluded as we believe this data to be associated with the response/dependent variable, crime, as opposed to a predictor/independent variable.

county - the variable county is indentified as the county identifer and we identify that each observation is unique.

```{r}
length(data$county) - length(unique(data$county))
```
Beyond identifying that each observation is unique, we may then dispense with this variable, as the ordinal number associated with each observation does not provide any useful information about the associated data.

From this first pairing down of variables, we are left with 19 variables of interest.  

```{r}
kept <- c(3,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,25)

data <- data[kept]
```


From these variables, we have grouped them into categories based upon their type of effect in our models: Authority, Demographics, Geography, and Economics.  Moving forwards, we will discuss variables based upon these groupings.

##Authority
Authority is ______.

We have identified is the police variable, described as "police per capita" from the code book.


##Demographics
We are interpereting demographics as _____.

There are two variables in the demographics category, ymale and pctmin, or proportion of county males between ages 15 and 24, and proportion that is minority or nonwhite, respectively.  

##Geography
Geography in our model represents _____.

There are a number of variables we have identified that relate to geography, including density (people per sq. mile), west and central, binary variables illustrating whether the observed county is in the west or central portion of the state, and urban, another binary variable representing if the county is in the standard metropolitan statistical area.

## Economics
Finally, economics is in our model as _____.

Included in this group of variables include all the weekly wage variables (wagecon (Construction), wagetuc (Transportation, Utilities, Communciations), wagetrd, (wholesale, retail trade), wagefir (finance, insurance and real estate), wagemfg (manufacturing), wagefed (Federal Employees), wagesta (state employees), and wageloc (local government employees)).  Finally we have also categorized tax, or the tax revenue per capita as an economics variable.


WARNING: Because we don't know the population of the counties or denominator of the average wages, the "per capita/averages" could be signficantly skewed or not reliably comparable across all counties.  We nevertheless proceed with caution.  


# Model Building
## Model 1: 
#### Specification

#### Assumptions

## Model 2:  
#### Specification

#### Assumptions

## Model 3:  
#### Specification

#### Assumptions

## Summary of Models

# Summary of models
## Regression Table

## Significance
### Statistical Significance

### Practical Significance



# Causality & Omitted Variables

# Conclusion
